Think again?\\ Optimal mental simulation tracks problem difficulty

### 2015-01-31 manuscript.tex
One sentence per line

Recent research in the domain of physical reasoning has revealed that people make predictions like these by running noisy physical simulations \cite{Smith:2013fc,Battaglia2013,Smith:2013ug,Smith:2013th,Smith:2014tx,Ullman:2014ut,Hamrick:2015} - the noisy Newton hypothesis \cite{Sanborn2013}. However, while this research has investigated the \emph{mechanism} by which people make these types of predictions, there has been very little investigation into \emph{how} people actually use this mechanism. If we are running mental simulations, how to we choose how many sample simulations to run?

In this paper, we focus on a dichotomous prediction task---will a ball in motion on a computer screen go through a hole or miss it? What would be considered an `efficient' mechanism for taking samples will depend on the structure of the task, and there is a large body of literature discussing how a sample-based agent should make a decision between two choices. Specifically, we consider the \textit{sequential probability ratio test}, or SPRT\cite{wald1947sequential} strategy, in which an agent takes samples that point to one hypothesis or another, and continues to do so until the net samples in favor of one hypothesis reaches a threshold, at which point that hypothesis wins. SPRT has been used successfully to explain behavior in a number of decision-making tasks \cite<e.g.>{Gold:2007fo,Ratcliff:2008ux,Bitzer:2014ea}, as it provides an optimal cost / benefit tradeoff between sampling and exploiting information \TODO{Can we claim this? If so, we need a citation}. While SPRT is traditionally applied to simple binary decisions (e.g., go left or go right), this framework can be extended with probabilistic physical simulations to make binary decisions about our world.

Crucially, the SPRT framework allows us to make testable predictions about how quickly people perform these dichotomous tasks. Because it involves sampling until the net evidence reaches a threshold, the number of samples will depend on the variability of the hypotheses. If samples always provide the same information, this net evidence will reach the threshold quickly, and fewer samples will be taken. On the other hand, if samples are more evenly distributed, it will take more samples to reach the net evidence threshold. Thus if people are using an SPRT strategy, we would expect that reaction times on a dichotomous task will vary systematically with task difficulty.

@@

\subsection{Selecting responses} Independent of the number of samples actually taken, the probability of choosing $H_1$ given that $H_1$ is the MAP hypothesis is:
\begin{equation}
\Pr[\hat{H}(Y_N)=H_1\,|\,H_1]=\frac{p^T}{p^T+(1-p)^T},
\label{eq:pr-choose-h1}
\end{equation}

\subsection{Expected number of samples} Given the threshold $T$, the probability of taking $N$ samples and then choosing $H_1$ is:
\begin{multline}
\Pr[N,H_1\,|\,T,p]=\left(\frac{2^N}{2T}\right)p^{\frac{N+T}{2}}(1-p)^{\frac{N-T}{2}}\cdot{}\\
\sum_{\nu=1}^{2T-1}\cos\left(\frac{\nu\pi}{2T}\right)^{N-1}\sin\left(\frac{\nu\pi}{2T}\right)\sin\left(\frac{\nu\pi}{2}\right)
\end{multline}
for $N\geq T$ and where $N$ is of the same parity as $T$ \cite[ch.~XIV, eq. 5.7]{Feller:1968ut}. Because we are dealing with binary hypotheses, the marginal probability of $N$ samples is then:
\begin{equation}
\Pr[N\,|\,T,p]=\Pr[N,H_1\,|\,T,p]+\Pr[N,H_1\,|\,T,1-p].
\label{eq:pr-n}
\end{equation}
and so the expected number of samples for a particular decision is:
\begin{equation}
\mathbb{E}[N\,|\,T,p]=\sum_{N=T}^\infty N\cdot{}\Pr[N\,|\,T,p]
\label{eq:expsamp}
\end{equation}

@@

On each trial, participants were shown the scene, including the initial position of the ball and the location of the hole. Participants were instructed to press the ``space'' key to begin the trial. Immediately upon pressing ``space'', the initial stimulus presentation began. As soon as the initial stimulus animation concluded, a gray box was drawn over the screen, occluding the ball (but not the line depicting the path it had traveled so far; this was left in as a reminder to participants of where the ball had come from). Participants were asked, ``will the ball go in the hole?'', and were instructed to press `q' to respond in the affirmative, and `p' to respond in the negative. Immediately after responding, text appeared saying ``Correct!'' or ``Incorrect.'', depending on the participants' response. Additionally, the gray occluder was removed, and participants were shown feedback animation of the path of the ball. After the feedback animation was complete, the final frame of the animation remained on the screen until participants pressed ``space'' to advance to the next trial.

@@

The stimuli consisted of animations depicting a blue ball with a radius of 10px bouncing around in a box with dimensions 900px $\times$ 650px. All stimuli consisted of two separate animations. The first was the stimulus presentation, which had a duration of 0.775 seconds and depicted the ball moving in a particular direction. The second animation was the feedback, which picked up immediately where the first animation left off, and which had a duration of 1.5 seconds and depicted the ball either going into the hole or bouncing off the wall that contained the hole.\footnote{To enforce the constraint that the ball always travel the same distance during the feedback animation, the $x$-coordinate of the wall with the hole in it was allowed to vary across stimuli.} During the feedback animation, the ball could bounce on the other walls either 0, 1, or 2 times before going into the hole or hitting the wall with the hole in it. In both animations, the ball had a velocity of 400px/s. Additionally, in the animations, the path that the ball had traveled so far was drawn with a faded gray line (see Figure \ref{fig:experiment}).

@@

We fit $\sigma_{adj}$ and $T$ using a grid search to maximize the correlation between SPRT response times (Equation \ref{eq:rt}) and actual response times in Experiment 1. To compute Equation \ref{eq:rt}, we ran $n=10000$ samples from the model, and from these samples (as well as the $\sigma_{adj}$ adjustment) estimated a truncated normal distribution along the wall with the hole in it. From this truncated normal, we computed the value of $p$ as the the probability density lying within the region of the hole. From this procedure, we found the best fitting parameters to be \sdadj{} and \threshold{}.\footnote{If \sdadj{}, then $M<1$. How could people be taking fewer than one sample? We suspect that $\sigma_{adj}<1$ because the model actually overestimates the standard deviation; thus, $\sigma_{adj}$ is adjusting for both the inflated uncertainty in the model as well increased certainty in people's judgments on the paddle task.}

@@

As in Experiment 1, immediately upon pressing ``space'', the initial stimulus presentation began, after which an occluder appeared. However, unlike Experiment 1, as soon as the occluder appeared, a timer also appeared and begin counting down for 2 seconds. During this time, participants had to move the paddle to try to catch the ball in the position it would be when the timer was up.When the timer countdown finished, the paddle froze at its current location, and the occluder was removed and the full path of the ball revealed. If the ball was on the paddle, then participants were told, ``You caught the ball!''. If the ball was not on the paddle, participants were told ``Oops, you didn't catch the ball.'' Participants were then instructed to press ``space'' to continue, at which point the next trial would begin.

@@

\subsection{Results}
We fit the model to people's responses in two steps. First, we fit $\sigma_p$, $\kappa_v$, $\kappa_m$, $\kappa_b$, and $\sigma_0$ to particant's responses in Experiment 2 \cite<for details, see>{Smith:2013fc}. Using this procedure, we found the best fitting parameters to be \perr{}, \kapv{}, \kapm{}, \kapb{}, and \sdzero{}. With these parameters, we find very similar results to those from \citeA{Smith:2013fc} (Figure \ref{fig:model-results}, left subplot). In particular, we found a correlation of \PaddleCorr{} between the model's predicted means of where the ball would end up and people's average location of the paddle.

>>>>

Recent research in the domain of physical reasoning has revealed that people make predictions like these by running noisy physical simulations \cite{Smith:2013fc,Battaglia2013,Smith:2013ug,Smith:2013th,Smith:2014tx,Ullman:2014ut,Hamrick:2015}---the noisy Newton hypothesis \cite{Sanborn2013}. 
However, while this research has investigated the \emph{mechanism} by which people make these types of predictions, there has been very little investigation into \emph{how} people actually use this mechanism. 
If we are running mental simulations, how to we choose how many simulations to run?

In this paper, we focus on a dichotomous prediction task---will a ball in motion on a computer screen go through a hole, or miss it? 
To model this task, we turn to the large body of literature discussing how a sample-based agent should make a decision between two choices. 
Specifically, we consider the \textit{sequential probability ratio test}, or SPRT, in which an agent takes samples that point to one hypothesis or another, and continues to do so until the net samples in favor of one hypothesis reaches a threshold, at which point that hypothesis wins \cite{wald1947sequential}. 
SPRT has been used successfully to explain behavior in a number of decision-making tasks \cite<e.g.>{Gold:2007fo,Ratcliff:2008ux,Bitzer:2014ea}, as it provides an optimal cost / benefit tradeoff between sampling and exploiting information.
\TODO{Can we claim this? If so, we need a citation}
While SPRT is traditionally applied to simple binary decisions (e.g., go left or go right), this framework can be extended with probabilistic physical simulations to make binary decisions about our world.

Crucially, the SPRT framework allows us to make testable predictions about how quickly people perform these dichotomous tasks. 
Because it involves sampling until the net evidence reaches a threshold, the number of samples will depend on the variability of the hypotheses. 
If samples always provide the same information, this net evidence will reach the threshold quickly, and fewer samples will be taken. 
On the other hand, if samples are more evenly distributed, it will take more samples to reach the net evidence threshold. 
Thus, if people are using an SPRT strategy, we would expect that reaction times on a dichotomous task will vary systematically with task difficulty.

@@

\subsection{Selecting responses}

Independent of the number of samples actually taken, the probability of choosing $H_1$ given that $H_1$ is the MAP hypothesis is:
\begin{equation}
\Pr[\hat{H}(Y_N)=H_1\,|\,H_1]=\frac{p^T}{p^T+(1-p)^T},
\label{eq:pr-choose-h1}
\end{equation}

\subsection{Expected number of samples}

Given the threshold $T$, the probability of taking $N$ samples and then choosing $H_1$ is:
\begin{multline}
\Pr[N,H_1\,|\,T,p]=\left(\frac{2^N}{2T}\right)p^{\frac{N+T}{2}}(1-p)^{\frac{N-T}{2}}\cdot{}\\
\sum_{\nu=1}^{2T-1}\cos\left(\frac{\nu\pi}{2T}\right)^{N-1}\sin\left(\frac{\nu\pi}{2T}\right)\sin\left(\frac{\nu\pi}{2}\right)
\end{multline}
for $N\geq T$ and where $N$ is of the same parity as $T$ \cite[ch.~XIV, eq. 5.7]{Feller:1968ut}. 
Because we are dealing with binary hypotheses, the marginal probability of $N$ samples is then:
\begin{equation}
\Pr[N\,|\,T,p]=\Pr[N,H_1\,|\,T,p]+\Pr[N,H_1\,|\,T,1-p].
\label{eq:pr-n}
\end{equation}
and so the expected number of samples for a particular decision is:
\begin{equation}
\mathbb{E}[N\,|\,T,p]=\sum_{N=T}^\infty N\cdot{}\Pr[N\,|\,T,p]
\label{eq:expsamp}
\end{equation}

@@

On each trial, participants were shown the scene, including the initial position of the ball and the location of the hole. 
Participants were instructed to press the ``space'' key to begin the trial. 
Immediately upon pressing ``space'', the initial stimulus presentation began. 
As soon as the initial stimulus animation concluded, a gray box was drawn over the screen, occluding the ball (but not the line depicting the path it had traveled so far; this was left in as a reminder to participants of where the ball had come from). 
Participants were asked, ``will the ball go in the hole?'', and were instructed to press `q' to respond in the affirmative, and `p' to respond in the negative. 
Immediately after responding, text appeared saying ``Correct!'' or ``Incorrect.'', depending on the participants' response.
Additionally, the gray occluder was removed, and participants were shown feedback animation of the path of the ball
After the feedback animation was complete, the final frame of the animation remained on the screen until participants pressed ``space'' to advance to the next trial.

@@

The stimuli consisted of animations depicting a blue ball with a radius of 10px bouncing around in a box with dimensions 900px $\times$ 650px.
All stimuli consisted of two separate animations.
The first was the stimulus presentation, which had a duration of 0.775 seconds and depicted the ball moving in a particular direction.
The second animation was the feedback, which picked up immediately where the first animation left off, and which had a duration of 1.5 seconds and depicted the ball either going into the hole or bouncing off the wall that contained the hole.
\footnote{To enforce the constraint that the ball always travel the same distance during the feedback animation, the $x$-coordinate of the wall with the hole in it was allowed to vary across stimuli.}
During the feedback animation, the ball could bounce on the other walls either 0, 1, or 2 times before going into the hole or hitting the wall with the hole in it.
In both animations, the ball had a velocity of 400px/s.
Additionally, in the animations, the path that the ball had traveled so far was drawn with a faded gray line (see Figure \ref{fig:experiment}).

@@

We fit $\sigma_{adj}$ and $T$ using a grid search to maximize the correlation between SPRT response times (Equation \ref{eq:rt}) and actual response times in Experiment 1.
To compute Equation \ref{eq:rt}, we ran $n=10000$ samples from the model, and from these samples (as well as the $\sigma_{adj}$ adjustment) estimated a truncated normal distribution along the wall with the hole in it.
From this truncated normal, we computed the value of $p$ as the the probability density lying within the region of the hole.
From this procedure, we found the best fitting parameters to be \sdadj{} and \threshold{}.
\footnote{If \sdadj{}, then $M<1$. How could people be taking fewer than one sample?
We suspect that $\sigma_{adj}<1$ because the model actually overestimates the standard deviation; thus, $\sigma_{adj}$ is adjusting for both the inflated uncertainty in the model as well increased certainty in people's judgments on the paddle task.}

@@

As in Experiment 1, immediately upon pressing ``space'', the initial stimulus presentation began, after which an occluder appeared.
However, unlike Experiment 1, as soon as the occluder appeared, a timer also appeared and begin counting down for 2 seconds.
During this time, participants had to move the paddle to try to catch the ball in the position it would be when the timer was up.
When the timer countdown finished, the paddle froze at its current location, and the occluder was removed and the full path of the ball revealed.
If the ball was on the paddle, then participants were told, ``You caught the ball!''.
If the ball was not on the paddle, participants were told ``Oops, you didn't catch the ball.''
Participants were then instructed to press ``space'' to continue, at which point the next trial would begin.

@@

We fit the model to people's responses in two steps. 
First, we fit $\sigma_p$, $\kappa_v$, $\kappa_m$, $\kappa_b$, and $\sigma_0$ to particant's responses in Experiment 2 \cite<for details, see>{Smith:2013fc}. 
Using this procedure, we found the best fitting parameters to be \perr{}, \kapv{}, \kapm{}, \kapb{}, and \sdzero{}. 
With these parameters, we find very similar results to those from \citeA{Smith:2013fc} (Figure \ref{fig:model-results}, left subplot).
In particular, we found a correlation of \PaddleCorr{} between the model's predicted means of where the ball would end up and people's average location of the paddle.
